Well-being of Individuals from the United States Global Social Survey are Affected by Finances, Relationships, and Health

### 2023-03-17 outputs/paper/paper3.qmd
Changed the tables, references, and added references

---
title: "Well-being of Individuals from the United States Global Social Survey are Affected by Finances, Relationships, and Health"
author: "Justin Abando, Jin Di Liu, Marcin Jaczynski"
thanks: "Code and data are available at: https://github.com/jackchinski/happiness_report"
date: "`r Sys.time()`"
date-format: "D MMMM YYYY"
abstract: "DON'T FORGET THIS"
format: pdf
bibliography: references.bib
---

@@

# 1 Introduction
##Literature Review
# 2 Data
This report focuses on the General Social Survey (GSS), which is a survey that observes and monitors public opinion and behavior in the United States ("US General Social Survey", 2021). Since 1972, the GSS has been conducted by the NORC at the University of Chicago, and is currently funded by the National Science Foundation (NSF) This survey was conducted annually until 1990, where it shifted towards a bi-annual schedule. Its aim is to maintain a consistent set of questions that aggregates American sentiments through similar sampling and questioning approaches. Depending on the year, political context, and contemporary academic practices, additional questions may be added or altered. However, there are still a core set of questions that are maintained and given every year. These core survey questions are mostly identity related (age, income, gender, etc.), but a few are also relevant to general opinions and well being. Additionally, the survey consists of repeating and topical modules, with each participant receiving a subset of both. For the scope of this study, we are going to look at survey responses produced from the 2021 survey.

## 2.1 Tools Used

## 2.2 Survey Methodology

@@

## 2.3 Survey Biases and Ethics
The lack of an interviewer implied the possibility of non-response, as participants are less socially pressured to fill out responses they may or may not feel inclined to report. To address this, an introductory form was administered to all participating adults which somewhat ensured that those taking the survey are dedicated enough to offer adequate responses.This was a trade off in the sampling approach that must be noted compared to previous years. Therefore, weights were carefully placed to account for factors such as population totals, density of surveyed addresses nearby, and non-response. The weighting system must account for new data and be similar to previous years to avoid skewing the results. However, despite all of these measures and countermeasures, the greatest challenge against the GSS (in all years) is the threat of survey bias, as all the data collected for all surveys are from participants willing to fill out a 90 minute survey. This could mean that the only people represented in the GSS survey are those with the free time and willingness to offer survey responses to NORC, potentially misrepresenting the overall population.

@@

## 2.4 Data Collection

In order to visualize the results of the GSS, the data was extracted from the GSS Data Explorer and was imported and analyzed using R \[\@rcoreteam2022\]. The tidyverse package \[\@wickham2019\] was used to load in the dplyr package to filter variables \[\@wickham2023\], and ggplot2 to visualize the data in graphs \[\@wickham2016\]. Furthermore, we also employed packages gtsummary and flextable to create table visualizations and summaries of our findings

@@

## 2.5 Data Cleaning

>>>>

---
title: "Well-being of Individuals from the United States Global Social Survey are Affected by Finances, Relationships, and Health"
author: "Justin Abando, Jin Di Liu, Marcin Jaczynski"
thanks: "Code and data are available at: https://github.com/jackchinski/happiness_report"
date: "`r Sys.time()`"
date-format: "D MMMM YYYY"
abstract: "This paper discusses data taken from the 2021 US General Social Survey to analyze the relationship between financial well being and overall individual well being. "
format: pdf
bibliography: references.bib
---

@@

# 1 Introduction

The advent of 2023 is one marked with economic hardship and depression. After the pandemic, markets scrambled as inflation rose to an all time high and signs of economic downturn began creeping through in the form of layoffs and reduced spending.
Because of this, we wanted to investigate the effects that finances (economic statuses and hardships) have on well being. For our investigation, we wished to interrogate elements that we believe are representative of well being in relation to financial wellbeing.  Specifically, we chose to investigate factors such as happiness, health, happiness of marriage, and life excitement. We were able to identify interesting correlations between quality of health, happiness, and happiness of marriage with financial satisfaction and income.

To first get a better grasp of this investigation, we conducted a small literature review on the relationships between finances and well being in the United States. A common well known relationship between well being and finances is the positive correlation between Life Satisfaction and Economic well being \[\@helliwell2012world\]. Furthermore, there are other economic theories that also attempt to interrogate the relationship between happiness and economic well being such as the Easterlin Paradox \[\@easterlin1974does\].
In this regard, the positive correlation between finances and happiness is well understood. However, looking deeper into the literature, we can also identify how the deprivation of economic well being can also negatively impact an individual’s quality of life. Specifically, research shows that economic hardship can impact marital quality and cause instability. This is due to an increase in occupational and financial stress that generate frustration, stress, anger and depression. \[\@conger1990linking\] Furthermore, economic hardships can also inflict long-term consequences on children’s psychological well being in adulthood. \[\@sobolewski2005economic\]. Finally, economic hardships play a major role in increasing depression levels, as depression can emerge as a byproduct of low income, education, being young, and having young children. \[\@ross1985hardship\]
	Based on our brief literature review and the current economic climate of 2023, we were inspired to engage with the GSS 2021 survey data to answer questions about the impact of financial satisfaction with overall well being.

# 2 Data
This report focuses on the General Social Survey (GSS), which is a survey that observes and monitors public opinion and behavior in the United States ("US General Social Survey", 2021). Since 1972, the GSS has been conducted by the NORC at the University of Chicago, and is currently funded by the National Science Foundation (NSF) This survey was conducted annually until 1990, where it shifted towards a bi-annual schedule. Its aim is to maintain a consistent set of questions that aggregates American sentiments through similar sampling and questioning approaches. Depending on the year, political context, and contemporary academic practices, additional questions may be added or altered. However, there are still a core set of questions that are maintained and given every year. These core survey questions are mostly identity related (age, income, gender, etc.), but a few are also relevant to general opinions and well being. Additionally, the survey consists of repeating and topical modules, with each participant receiving a subset of both. For the scope of this study, we are going to look at survey responses produced from the 2021 survey.

## 2.1 Survey Methodology

@@

## 2.2 Survey Biases and Ethics
The lack of an interviewer implied the possibility of non-response, as participants are less socially pressured to fill out responses they may or may not feel inclined to report. To address this, an introductory form was administered to all participating adults which somewhat ensured that those taking the survey are dedicated enough to offer adequate responses.This was a trade off in the sampling approach that must be noted compared to previous years. Therefore, weights were carefully placed to account for factors such as population totals, density of surveyed addresses nearby, and non-response. The weighting system must account for new data and be similar to previous years to avoid skewing the results. However, despite all of these measures and countermeasures, the greatest challenge against the GSS (in all years) is the threat of survey bias, as all the data collected for all surveys are from participants willing to fill out a 90 minute survey. This could mean that the only people represented in the GSS survey are those with the free time and willingness to offer survey responses to NORC, potentially misrepresenting the overall population.

@@

## 2.3 Data Collection

In order to visualize the results of the GSS, the data was extracted from the GSS Data Explorer and was imported and analyzed using R \[\@rcoreteam2022\]. The tidyverse package \[\@wickham2019\] was used to load in the dplyr package to filter variables \[\@wickham2023\], and ggplot2 to visualize the data in graphs \[\@wickham2016\]. Furthermore, we also employed packages gtsummary \[\@gtsummary\] and \[\@flextable\]  to create table visualizations and summaries of our findings

@@

## 2.4 Data Cleaning
